Dark Control:\\ A Unified Account of Default Mode Function\\ by Control Theory and Reinforcement Learning

### 2017-06-10 paper/paper.tex
fixing Fig 3 caption layout
pass over FEP

\begin{wrapfigure}{r}{.4\linewidth}
  \caption{\textbf{Reinforcement learning in a nutshell.} Given the current state of the environment,
    the agent takes an action by following some policy, receives a consequential reward and observes the next state. The process goes on until interrupted or a goal state is reached.}
  \centering
  \includegraphics[width=1.\linewidth]{rl.pdf}
  \label{fig:rl}
\end{wrapfigure}
The environment is generally taken as \textit{stochastic},
that is, changing in random ways. In addition, the environment is only
\textit{partially observable} in the sense that only limited aspects of the environment's
state are accessible to the agent's sensors.
\citep{starkweather2017dopamine}.
We assume that volatility of the environment
is realistic in a computational model which sets out
to explain DMN functions of the human brain.
% XXX: the following passage is already in the introduction
% Indeed, it is known that the more the external world is predictable,
% the more mental activity becomes detached from the actual sensory inputs
% \citep{antrobus1966studies, pope1978regulation},
% the more stimulus-independent thoughts occur, and
% the higher the neural activity in the DMN.
% Conversely, the more the currently executed task is unknown and unpracticed,
% the lower the DMN activity
% \citep{filler1973daydreaming, teasdale1995stimulus}.
% %
% These ``offline'' processes may however contribute to
% optimizing control of the organism.
% Such a framework of DMN funtion based on reinforcement learning
% can naturally embed human behavior
% in the tension between exploitative action with immediate gains and
% explorative action with longer-term reward schedules
% \citep{dayan2008decision}.
We argue that a functional account of the DMN based on RL
can naturally embed human behavior
in the tension between exploitative action with immediate gains and
explorative action with longer-term reward outcomes
\citep{dayan2008decision}.

@@

\subsection{The free-energy principle and active inference}
In Friston's free-energy principle (FEP) \citep{friston2010free} and active inference  (i.e action under the FEP) \citep{fristonAIorRL} theories, the brain is portrayed
as biomechanical inference engine
dedicated to minimizing the long-term average of surprise: the log-likelihood of the observed sensory input --or more precisely, an upper bound thereof-- \textit{relative} to the expectations of, and the internal representations the brain has about the external world. It does this by actively generating hypothetical explanations or predictions of its sensory input.
Precursors of this theory can be traced back to \citep{dayan1995helmholtz} in which they introduced the so-called \textit{Helmholtz machine}, a hierarchical factorial \textit{directional deep belief-net}. According to the FEP account,
the goal of the brain is then to optimize over a generative model $G$ of sensations: to iteratively
modify its internal representation $p_G(\z|\x)$ about objects in the world, their interactions and dynamics,
so as to minimize surprise when these representations are confronted with sensory input $\x$ during perception cycles (i.e., the \textit{generative} model).
The FEP also postulates a dual model called the \textit{recognition} model,
which works in tandem with the generative model $p_R(\z|\x)$, to accomplish approximate inference. The recognition model dreams imaginary worlds $\z$,
while the latter tries to generate sensations $\x$ which match these dreams!

Because surprise is intractably difficult to optimize (since we would need to sum over all hidden causes of the sensations), FEP sets out to instead minimize an upper-bound thereof,
namely the \text{free-energy} $F_G^R(\x)$ given by

@@

\subsubsection{Criticisms of the FEP}
  Despite its popularity, a number of criticisms against the FEP have surfaced over the years. We outline the most notable ones here.

\paragraph{The wake-sleep algorithm is not neurobiologically plausible.}
The main algorithm for minimizing free-energy $F_G^R(\x)$ is the \textit{wake-sleep algorithm}
\citep{dayan1995helmholtz}. As noted already in \citep{dayan1995helmholtz},
a crucial drawback of the wake-sleep algorithm is that it postulates a pair forward (generation) and backward (recognition) models,
which together do not correspond to optimization of (a bound of) the marginal likelihood, because of the incorrect KL used therein, etc.
Thus the brain could not possibly be running such a algorithm, not even in principle! To the rescue, we note that the recent theory of
\textit{variational auto-encoders (VAEs)} \citep{kingma2013auto} might provide an efficient alternative to the wake-sleep algorithm,
as it overcomes the technical limits of the former, by using a reparametrization trick. For instance, unlike the wake-sleep algorithm
for minimizing free-energy, VAEs can be efficiently trained via back-propagation of prediction errors.

\paragraph{The ``dark room'' experiment.}
Since FEP-based theories \citep{friston2010free,fristonAIorRL} postulate that all ongoing intelligent behavior in an organism is geared towards reaching the grand goal of minimizing surprisal, an organism which enters a \textit{dark room}\footnote{\textit{Dark} is used here as an allegory to indicate the complete absence of any \textit{external} sensory input.} (Fig. \ref{fig:darkroom}) would strive to remain there as its sensory input would be perfectly predictable in such a state \citep{darkroom2012}.
However, such a tendency is definitely not what is observed, neither in humans nor in animals: if you get into a dark room,
\begin{wrapfigure}{r}{.45\linewidth}
  \centering
  \includegraphics[width=1.\linewidth]{darkroom.jpg}
  \caption{\textbf{The \textit{dark room} experiment.} The adjective \textit{dark} is used as an allegory to indicate the complete absence of any \textit{external} sensory input.}
  \label{fig:darkroom}
\end{wrapfigure}
you try to figure out where the light switch is, so as to \textit{explore} this mysterious room, or simply get out of it! Defenders of the FEP have retaliated by advancing the `full package''\citep{darkroom2012}: FEP is multi-scale, and there would be a meta-scale at which the organism would
be \textit{surprised} by such a \textit{lack of surprise}. According to this argument, a dark room is thus paradoxically, a state of extremely high surprise. Driven by the zeal to always diminish surprise, the FEP agent would eventually bootstrap itself out of such saddles, to explore more interesting corners of the environment. Recourse to this argument is equivalent to invoking a \textit{halting oracle}\citep{haltingoracle}, an arbitrarily powerful machine which can solve the \textit{halting problem} (an arbitrarily difficult computational problem which cannot be solved under any reasonable \textit{model of computation}, not even in principle!). But this is problematic as it sucks up the problem (that of decision-making and active agency) the theory intends to elude in the first place.

In contrast, operating under our RL-based theory, an organism would quickly notice that this dark room is a local minimum. Indeed, hippocampal experience replay (see \ref{sec:self}) would set it and the organism would sample memories or fantasies of alternative situations in which there was reward to mine. Such considerations would artificially generate \textit{internal} sensory input for the DMN and eventually let the organism to explore the room, for example look for and find the light switch, or simply leave the room.

\subsubsection{RL reinterpretation of FEP and active inference}  FEP and active inference can be reinterpreted in our model-free RL framework by postulating that the reward signal maximized by the DMN is the negative of surprisal, i.e the log-likehood of sensory the current priors the agent has about the world. This is worth writing down explicitly,

\begin{equation}
  r^{\text{FEP}} = \underbrace{\log (p_G)}_{-\text{FEP generative surprisal}}.
    \end{equation}
% \paragraph{Comparison to our proposed theory.}
% On the surface, a common point between the FEP and our proposed RL-based framework
%   is placing the minimization of a surprise signal at the core of brain function.
%   Indeed in RL, surprise minimization is subsumed  by accurate prediction of
%   rewarding outcomes in the future. A ``free-energy'' agent is barely a biomechanical machine which has the tendency to resist undesired phase-transitions
%   \citep{friston2010free,fristonAIorRL,ortega2013thermodynamics}.
    Such a surprisal-based reinforcement learning scheme has been advocated for in \citep{schmidhuber2010,mohamed2015} under the equivalent framework of information compression.
    However, we note that though relevant, minimization of surprisal alone is insufficient to explain intelligent behavior observed in humans and other ``intelligent'' animals. The dark-room experiment, for example, provides a formidable argument against in favor of the insufficiency of the FEP and active inference theories to explain intelligent behavior in humans: when a human is put in a dark room, they don't just sit down in a corner forever, what would be the FEP-optimal thing to do; they actively search for the light switch, and start exploring the depths of the room, possibly looking for food or the way out!

>>>>

\begin{wrapfigure}{r}{.3\linewidth}
  \caption{\textbf{Reinforcement learning in a nutshell.} Given the current state of the environment,
    the agent takes an action by following some policy, receives a consequential reward and observes the next state. The process goes on until interrupted or a goal state is reached.}
  \centering
  \includegraphics[width=1.\linewidth]{rl.pdf}
  \label{fig:rl}
\end{wrapfigure}
The environment is generally taken as \textit{stochastic},
that is, changing in random ways. In addition, the environment is only
\textit{partially observable} in the sense that only limited aspects of the environment's
state are accessible to the agent's sensors.
\citep{starkweather2017dopamine}.
We assume that volatility of the environment
is realistic in a computational model which sets out
to explain DMN functions of the human brain.
% XXX: the following passage is already in the introduction
% Indeed, it is known that the more the external world is predictable,
% the more mental activity becomes detached from the actual sensory inputs
% \citep{antrobus1966studies, pope1978regulation},
% the more stimulus-independent thoughts occur, and
% the higher the neural activity in the DMN.
% Conversely, the more the currently executed task is unknown and unpracticed,
% the lower the DMN activity
% \citep{filler1973daydreaming, teasdale1995stimulus}.
% %
% These ``offline'' processes may however contribute to
% optimizing control of the organism.
% Such a framework of DMN funtion based on reinforcement learning
% can naturally embed human behavior
% in the tension between exploitative action with immediate gains and
% explorative action with longer-term reward schedules
% \citep{dayan2008decision}.
We argue that a functional account of the DMN based on RL
can naturally embed human behavior
in the tension between exploitative action with immediate gains and
explorative action with longer-term reward outcomes
\citep{dayan2008decision}.

@@

\subsection{The free-energy principle and active inference}
% INTRO
According to the free-energy principle (FEP) and
active inference theories \citep{friston2010free, fristonAIorRL},
the brain corresponds to
a biomechanical inference engine. It is
dedicated to minimizing the long-term average of surprise: the log-likelihood of the observed sensory input --or more precisely, an upper bound thereof-- \textit{relative} to the expectations about the external world derived from the brain's internal representations. The brain
would continuously generate hypothetical explanations of the world
and predict its sensory input.
Precursors of this theory can be traced back to Dayan and colleagues \citep{dayan1995helmholtz}
who introduced the \textit{Helmholtz machine}, a hierarchical factorial directional deep belief network. According to the FEP account,
the goal of the brain is to optimize over a generative model $G$ of sensations: to iteratively
modify its internal representation $p_G(\z|\x)$ about objects in the world, their interactions and dynamics. The model then allows
to minimize surprise when these representations are confronted with sensory input $\x$ during action-perception cycles (i.e., the \textit{generative} model).
The FEP also postulates a dual model called the \textit{recognition} model,
which works in tandem with the generative model $p_R(\z|\x)$ to accomplish approximate inference. Put differently, the recognition model dreams imaginary worlds $\z$,
while the generative model tries to produce sensations $\x$ which match these
imagined mental scenarios.
Howewer, surprise is challenging to optimize numerically
because we need to sum over all hidden causes of the sensations.
Instead, FEP therefore minimizes an upper-bound of surprise,
namely the \text{free-energy} $F_G^R(\x)$ given by

@@

% CRITICISM
Despite its popularity, criticism against the FEP has arisen over the years,
some of which we outline in the following.
The main algorithm for minimizing free-energy $F_G^R(\x)$ is the \textit{wake-sleep algorithm}
\citep{dayan1995helmholtz}. As previously noted \citep{dayan1995helmholtz},
a crucial drawback of the wake-sleep algorithm is that it postulates a pair
of forward (generation) and backward (recognition) models that
together do not correspond to optimization of (a bound of) the marginal likelihood.
Thus, the brain is unlikely to implement a variant of the wake-sleep algorithm.
We note that the recent theory of
\textit{variational auto-encoders} (VAEs) \citep{kingma2013auto} might provide an efficient alternative to the wake-sleep algorithm. This is because VAEs
overcome a number of the technical limits of the wake-sleep algorithm
by using a reparametrization trick. For instance, unlike the wake-sleep algorithm
for minimizing free-energy, VAEs can be efficiently trained via back-propagation of prediction errors.

Since FEP-based theories \citep{friston2010free,fristonAIorRL} conceptualize ongoing behavior
in an organism to be geared towards the surprise-minimizing goal,
an organism entering a dark room
% \footnote{\textit{Dark} is used here as an allegory to indicate the complete absence of any \textit{external} sensory input.}
(Fig. \ref{fig:darkroom}) would strive to remain in this location
because its sensory inputs are perfectly predictable
in this environmental state \citep{darkroom2012}.
However, such a behavioral tendency is
seldom observed in animals or humans in the real world: In a dark room,
\begin{wrapfigure}{r}{.45\linewidth}
  \centering
  \includegraphics[width=1.\linewidth]{darkroom.jpg}
  \caption{\textbf{The dark room experiment.} An intelligent agent situated
  in a closed space is used as a thought experiment for the complete absence
  of external sensory input.}
  \label{fig:darkroom}
\end{wrapfigure}
intelligent agents would search for light sources to \textit{explore}
its surroundings or leave it.
Defenders of the FEP have retorted by advancing the ``full package'' \citep{darkroom2012}:
FEP is proposed to be multi-scale and there would be a meta-scale at which the organism would
be \textit{surprised} by such a \textit{lack of surprise}. According to this argument, a dark room would paradoxically correspond to a state of particularly high surprise.
Driven by surprise-minimization objective, the FEP agent would eventually bootstrap itself out of such saddles to explore more interesting parts of the environment.
% Recourse to this argument is equivalent to invoking a \textit{halting oracle}\citep{haltingoracle}, an arbitrarily powerful machine which can solve the \textit{halting problem} (an arbitrarily difficult computational problem which cannot be solved under any reasonable \textit{model of computation}, not even in principle!). But this is problematic as it sucks up the problem (that of decision-making and active agency) the theory intends to elude in the first place.
In contrast, an organism operating under our RL-based theory would readily
identify the visual-input-free room as a local minimum. Indeed, hippocampal experience replay (see \ref{sec:self}) would serve to sample memories or fantasies of alternative situations
with reward structure. Such artificially generated \textit{internal} sensory input
subserved by the DMN can entice the organism to explore the room,
for instance by looking for and using the light switch or simply leaving the room.

Finally, we note that FEP and active inference can be recast in our model-free RL framework.
This becomes possible by postulating that the reward signal maximized by the DMN is
the negative surprise,
that is, the log-likehood of sensory the current priors the agent has about the world. More explicitly, this corresponds to

\begin{equation}
  r^{\text{FEP}} = \underbrace{\log (p_G)}_{-\text{FEP generative surprisal}}.
    \end{equation}
% \paragraph{Comparison to our proposed theory.}
% On the surface, a common point between the FEP and our proposed RL-based framework
%   is placing the minimization of a surprise signal at the core of brain function.
%   Indeed in RL, surprise minimization is subsumed  by accurate prediction of
%   rewarding outcomes in the future. A ``free-energy'' agent is barely a biomechanical machine which has the tendency to resist undesired phase-transitions
%   \citep{friston2010free,fristonAIorRL,ortega2013thermodynamics}.
Such a surprisal-based reinforcement learning scheme has previously been advocated \citep{schmidhuber2010,mohamed2015} under the equivalent framework of information compression.
Nevertheless, minimization of surprise alone
may be insufficient to explain behavior observed in humans and other intelligent animals.
% The dark-room experiment, for example, provides a formidable argument against in favor of the insufficiency of the FEP and active inference theories to explain intelligent behavior in humans: when a human is put in a dark room, they don't just sit down in a corner forever, what would be the FEP-optimal thing to do; they actively search for the light switch, and start exploring the depths of the room, possibly looking for food or the way out!
